Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-16 forcomparison.tex
Many graphs added and comments added. Edits of Ch.5. Starting to prep to add Grammarly back in

Experimenting on biological, physical, and artificial systems in order to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or on intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However our world around is, for the most part, much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they even provide us with a good physical intuition of the underlying principles governing the dynamics. 
To complicate this - the underlying systems we observe often display sensitive dependence on initial conditions; despite having highly similar initial conditions, differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually 'negligible' computational noise or measurement error renders long-term point-wise prediction infeasible. 
There are many difficulties that one encounters while modelling such systems:
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The actual system may be described by functions that behave wildly in the sense their graphs have a wild oscillatory behaviour, i.e., they have a large functional complexity.    
\end{enumerate}

@@

Examples in the literature attempting to forecast data from such systems have tried a number of different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

Recent partially interpretable models available in the literature have been based on the Koopman operator (e.g.,\cite{koopman1932dynamical,budivsic2012applied}) to  employ observables mapping the data onto a higher-dimensional space. This then makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods, obviously, do not guarantee an exact reconstructions for nonlinear models, and in practice provide poor long-term consistency for a large class of systems. \textbf{(Citation)}

The non-interpretable models include the delay embedding and the machine learning algorithms. One could learn a system conjugate to the underlying system by applying the Takens delay embedding \cite{takens1981detecting} when one has ``good" observations from the system; this learnt system could then be used for forecast the observed data.  Takens delay embedding theorem \cite{takens1981detecting} and its various generalisations (e.g., \cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then establishes the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality of this representation still depends on numerous parameters, making the comprehension of the dynamics rather unreliable (e.g., \cite{principe1992prediction}). One reason for this fragility is that the embedded attractor in the reconstruction space not always an attractor of the map that is learnt in the reconstruction space, despite  unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter \ref{ch3}) it can cause predictions to fail.

Practically, the application of Takens embedding involves learning a map through some technique and consequently one wishes that these would have low functional complexity\cite{manjunath2021universal}, i.e., functions with fewer oscillatory graphs. Pure machine learning methodology that processes temporal information (like the echo state networks (Citations to echo state networks)) by mapping data onto a higher dimensional space for further processing. Although they perform  well on forecasting some dynamical systems, they fail completely on others (Citations here) as there is often no guarantee that the right function was learnt during training.

This thesis deals with the implementation and analysis of non-interpretable models that can guarantee exact reconstruction. The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter 4). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems. 
\newline In Chapter~\ref{ch3}, we introduce the problem of forecasting dynamical systems, state the Takens delay embedding theorem and discuss various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

@@

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this thesis, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. 
Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are known at all times, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. 
In ecology, discrete-time dynamical systems are widely used to model population growth. The model in this case is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

@@

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$ \cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a larger system capturing all of the important dynamics of its factor.

@@

However, the true state $u$ of a system is seldom, if ever, fully known. In almost all cases we can at most insert probes into a system to obtain partial information by means of the measurements taken. Moreover, the process of taking a measurement itself introduces 2 additional aspects which complicate the problem: \ednote{B: Perhaps this edit makes more sense? M: Maybe replace difficulty by saying that we need to consider two aspects. }
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time-interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be fully accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

% The Takens embedding theorem does not guarantee global dissipativity.
% \ednote{B:Links with~\ref{subs_LearnGamma} in discussion of dissipativity of $\Gamma$.  Sir, I'm not sure how to write this here - I think it a beneficial point to make so as to set the stage for the later discussion in ~\ref{subs_LearnGamma}, but I am unsure there as well.}

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to being surjective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The continuity of $H^{-1}$ follows from the fact that $H$ is itself a continuous and surjective mapping of a compact space $\overleftarrow{U}$ in a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}. 
\ednote{B: edited the sentence here}

@@

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: Added so that FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: Note! We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

Our implementation of the reservoir(A) and input(B) matrices in the RNN makes use of sparse matrices. We discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opted to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed. \textbf{(Cite.)} By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input. In so doing we then also guarantee a measure of stability. \ednote{B: Not satisfactory. Edit again.}

>>>>

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually' negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. There are many difficulties that one encounters while modelling such systems:

\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

@@

Examples in the literature attempting to forecast data from such systems have tried many different approaches with differing degrees of success. When we have complete access to states of the system, an ordinary differential equation can be obtained from data ([?, ?] and [?, ?]) wherein one could approximate the vector field by a library of functions to get interpretable models.
Examples in the literature attempting to forecast data from such systems have tried a number of different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

Recent partially interpretable models available in the literature have been based on the Koopman operator (e.g.,[?, ?]) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice, provide poor long-term consistency for a large class of systems. (Citation)

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding [?] when one has "good" observations from the system; this learnt system could then be used to forecast the observed data. Takens delay embedding theorem [?] and its various generalisations (e.g., [?, ?, ?]) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (e.g., [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter 3), it can cause predictions to fail.

Practically, the application of Takens embedding involves learning a map through some technique, and consequently, one wishes that these would have low functional complexity[?], i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks (Citations to echo state networks)) by mapping data onto a higher dimensional space for further processing. Although they perform well on forecasting some dynamical systems, they fail on others (Citations here) as there is often no guarantee that the correct function was learnt during training.

This thesis involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. The project work concerns the study and implementation of a method ([?]) that incorporates learning a function by mapping the data onto a higher dimensional space using what is called a driven dynamical system (See Chapter 4). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows:
\newline In Chapter 2, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE, and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems.
\newline Chapter 3 introduces the problem of forecasting dynamical systems, states the Takens delay embedding theorem, and discusses various issues faced while forecasting.
\newline In Chapter 4, a driven dynamical system is defined and discusses the properties of a specific class of driven dynamical systems that we use in this thesis/report/paper.
\newline Finally, Chapter 5 shows the implementation of these forecasting methods, and conclusions are provided in Chapter 6.

@@

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this thesis, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are known at all times, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. In ecology, discrete-time dynamical systems are widely used to model population growth. The model, in this case, is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

@@

\ednote{M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly. B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough? }
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$ \cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a larger system capturing all of the important dynamics of its factor.

@@

However, the true state $u$ of a system is seldom, if ever, fully known. In almost all cases we can at most insert probes into a system to obtain partial information by means of the measurements taken. Moreover, the process of taking a measurement itself introduces 2 additional aspects which complicate the problem: \ednote{M: Maybe replace difficulty by saying that we need to consider two aspects. B: Perhaps this edit makes more sense?}
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time-interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be fully accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

The Takens embedding theorem does not guarantee global dissipativity.
\ednote{B:Links with~\ref{subs_LearnGamma} in discussion of dissipativity of $\Gamma$.  Sir, I'm not sure how to write this here - I think it a beneficial point to make so as to set the stage for the later discussion in ~\ref{subs_LearnGamma}, but I am unsure there as well.}

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to being surjective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The continuity of $H^{-1}$ follows from the fact that $H$ is itself a continuous and surjective mapping of a compact space $\overleftarrow{U}$ in a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}. 
\ednote{B: edited here}

@@

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: So that FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: Note! We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

Our implementation of the reservoir(A) and input(B) matrices in the RNN makes use of sparse matrices. We discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opted to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed. \textbf{(Cite.)} By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input. In so doing we then also guarantee a measure of stability. /ednote{B: Not satisfactory. Edit.}
